The effect of genetic linkage on island endemics richness

### 2018-08-24 paper/draft.tex
include JSC comments

\section{Introduction}

@@

In fact, these factors indeed often correlate strongly with the timing of genetic divergence of species and populations \cite{hembryEvolutionaryBiogeographyTerrestrial}. % REFS!
But even then, in some cases the reason for maintenance of population diversity remains in the dark. \cite{chenConcordanceGeneticDiversity}

@@

Additionally, these studies highlight the cases of successful radiations, while a conclusive explanation for why these population variablities and thus radiations could occur in the first place is still missing.

@@

However, the problem of investigating speciation processes is the time frame and study extent needed to go beyond mere correlation \cite{dormannCorrelationProcessSpecies}.
Because island biogeography processes and thus radiations generally take place at evolutionary time frames (millions of years),
comprehensive and meaningful data is difficult to obtain (\cite{didierLikelihoodTreeTopologies2017}, but see \cite{mitchellInferringDiversificationRate})
Therefore, there has been an interest in building process-based island models to test the causality of assumptions and hypotheses.
In a recent review, we indeed found a rapid increase in number of mechanistic simulation models for island biogeography \cite{leidingerBiodiversityDynamicsIslands2017}.
Although many of the reviewed models integrate multiple processes at a time, unfortunately,
most models failed to include micro-evolutionary or more explicit genetic processes %\cite{leidinger2017biodiversity}
(Leidinger \& Sarmento Cabral 2017).
One direction towards this is given by models using concepts such as ``protracted speciaton'' \cite{rosindellUnifiedModelSpecies2013}.
where speciation occurs with a delay from the actual
divergence event if the population remains isolated during that time.
But nature is usually more complex than this (cf.\ \cite{silvertownGhostCompetitionPhylogeny2004} but see \cite{herbenGhostHybridizationNiche2005}) 
and we expect genetic mechanisms to play a central role in speciation/radiation and species richness.
Questions, such as "What functional traits (e.g. relating to dispersal capacity,
reproduction, trophic ecology) are associated with high diversification rates within and across island systems?" %better another...
or "What is the influence of gene flow among islands and/or
between islands and mainland areas on speciation rates?", thus remain largely open \cite{patinoRoadmapIslandBiology2017}
and may require explicit consideration of evolutionary processes.
%% Yet, there are numerous examples for fascinating patterns of trait distributions and evolution on islands, such as island syndromes[, ...] or the fact of explosively radiating lineages (Hughes \& Eastwood 2006, Kocher 2004)
%%, while others remain one single species (Price/Chase) (eg. \cite{price2004speciation}).
%% Also: Ecological Opportunity and Adaptive Radiation, Stroud \& Losos 2016: ``Does the concept have predictive value[...]?'' \cite{stroud2016ecological}
%% Phylogenetic niche conservatism, phylogenetic signal and the relationship between phylogenetic relatedness and ecological similarity among species, Losos 2008: ecological vs. phylogenetic similarity \cite{losos2008phylogenetic}
Still, the most mechanisms integrated in island models [?] so far include only ecological
or geo-morphological processes (Rundell \& Price 2009, Esselstyn et al. 2011),
while genetic properties are hardly recognized in the island context.
Given the essential role islands have played in the development of evolutionary theories, it is surprising that these processes
are not implemented more often in island simulations.

@@

%% Aim, objectives:
%% In this study, we aim to address...
To assess the role of genetic processes, namely genetic linkage, for island radiations,
we present a mechanistic model that overcomes above-mentioned shortcomings like missing micro-evolutinary processes and lack of integrative design
by implementing genetic processes on a niche-explicit background
using an island biogeography simulation arena.
We will address the questions of what relative impact does (a) genetic linkage and
(b) reproductive isolation have on emerging species richness and composition,
and (c) which genetic and functional trait syndromes %% Which?
distinguish successful colonizer,
strong competitor and diversifying lineages.
Considering these phenomena, we expect speciation rates to be highest at intermediate levels of gene linkage --- low
levels leads to dissolving phenotype populations, whereas high levels hinder rapid adaption. %what Schiffers et al say about linkage and evolutionary rescue (basically the opposite of diversification?)

@@

\begin{figure}
  %\includegraphics{}
  \caption{Schematic of the model's individuals genetic architecture.
    Individuals are composed of a diploid set of linkage units which combine the genes coding for functional traits.}
  \label{schematic}
\end{figure}

We use an individual-based spatially and temporally explicit metacommunity model written in julia \cite{bezansonJuliaFreshApproach2017}.
The model considers competition between individuals, environmental niche adaptation, density independent mortality,
dispersal, sexual reproduction and mutation.
Individuals are characterized by a lineage identifier,
body mass, age and fitness.
Each individual's life history traits (table) are coded by mutable genes \cref{fig:schematic}.
The genes, in turn, may be combined to linkage units (``chromosomes'').
Genes in a linkage unit are considered a heredetary unit.
This means that they will always be inherited in junction.
Thus, only complete linkage units will be subject to recombination.

@@

\subsection{scenarios}
In the simulations, we modify the value and extent of
%(a)
gene linkage
%(b) reproductive isolation (or reproduction compatibility),
%(c) habitat diversity, and
%(d) topographical complexity
to asses their relative contribution on community composition.
Additionally, we simulate lineages in isolation with varying degrees of genetic linkage
and analyize emerging species numbers.
\cref{tab:scenarios} summarizes the parameters which were varied for the scenarios, their meaning and their values. %TODO: make table.
Each parameter combination was replicated %TODO: how much?
times.%, setting a different random seed for each replicate.

@@

@@

\begin{figure}
  %\includegraphics{}
  \caption{[2 x 2]
    Emerging numbers of species (and average population sizes) over time.
    Low-linkage lineages in COLOR1, high-linkage lineages in COLOR2.
    Top row: lineages simulated in isolation.
    Bottom row: simulated communities.}
  \label{specieslinkage}
\end{figure}

\begin{figure}
  %\includegraphics{}
  \caption{[2 x 2]
    Extinctions and numbers of species per lineage after XXXX timesteps against degree of genetic linkage.
    Top row: lineages simulated in isolation.
    Bottom row: simulated communities.}
  \label{specieslinkage}
\end{figure}

\begin{figure}
  %\includegraphics{}
  \caption{Trait space characteristics of lineages.
    (A) [Ordination] Trait space of successful colonizers compared to the mainland species pool. Each point represents a lineage.
    Arrows denote the traits.
    (B) [Ordination?] Linkage of traits of successful colonizers(/radiators?). Traits perpendicular to each other are inherited independently,
  same direction indicate tight linkage of traits.}
  \label{traitspace}
\end{figure}

@@

\section{Discussion}
The fact that low-linkage (single) species speciated in general faster than their high-linkage counterparts
is in line with expectations given their higher potential for fast evolutionary response [as exemplified REFS].
What did come as a surprise, however was that, over time, high linkage lineages surpassed low linkage lineages
in the number of emerging species.
The reason for this is likely to be found in the high genetic variation throughout the high-linkage populations
due to genetic hitchhiking.
Combined with random drift, this genetic variation has a high potential of producing numerous species over time [non-adaptive radiation].
[EXAMPLES]
On the other hand, independent recombination of genes in low-linkage populations explains their quick adaptive response
to the conditions in their respective environments.
Once adapted and having reached a high degree of specialization, however, the genetic architecture leads to strong selection against
new and mainly deleterious mutations.
[EXAMPLES]

@@

>>>>

\section{Introduction}
%% WHY ARE GENERA DIFFERENTLY RICH BESIDE ECOLOGY AND GEOGRAPHY/GEOLOGY? GENETICS!

@@

These factors indeed often correlate strongly with the timing of genetic divergence of species and populations \cite{hembryEvolutionaryBiogeographyTerrestrial}. % REFS!
But even then, in some cases the reason for maintenance of population variability remains unclear. \cite{chenConcordanceGeneticDiversity}

@@

Additionally, these studies highlight the cases of successful radiations, while a conclusive explanation for why these population variabilities and thus radiations could occur in the first place is still missing.

@@

The problem of investigating speciation processes is the time frame and study extent needed to go beyond mere correlation \cite{dormannCorrelationProcessSpecies}.
Because island biogeography processes and thus radiations generally take place at evolutionary time frames (millions of years),
comprehensive and meaningful data is difficult to obtain (\cite{didierLikelihoodTreeTopologies2017}, but see \cite{mitchellInferringDiversificationRate})
Therefore, there has been an interest in building process-based island models to test the causality of assumptions and hypotheses.
In a recent review, we indeed found a rapid increase in number of mechanistic simulation models for island biogeography \cite{leidingerBiodiversityDynamicsIslands2017}.
Although many of the reviewed models integrate multiple processes at a time, unfortunately,
most models failed to include micro-evolutionary or more explicit genetic processes %\cite{leidinger2017biodiversity}
(Leidinger \& Sarmento Cabral 2017).
One direction towards this is given by models using concepts such as ``protracted speciaton'' \cite{rosindellUnifiedModelSpecies2013}.
where speciation occurs with a delay from the actual
divergence event if the population remains isolated during that time.
But nature is usually more complex than this %(cf.\ \cite{silvertownGhostCompetitionPhylogeny2004} but see \cite{herbenGhostHybridizationNiche2005}) 
and we expect genetic mechanisms to play a central role in speciation/radiation and species richness.
Questions, such as "What functional traits (e.g. relating to dispersal capacity,
reproduction, trophic ecology) are associated with high diversification rates within and across island systems?" %better another...
or "What is the influence of gene flow among islands and/or
between islands and mainland areas on speciation rates?", thus remain largely open \cite{patinoRoadmapIslandBiology2017}
and may require explicit consideration of evolutionary processes.
%% Yet, there are numerous examples for fascinating patterns of trait distributions and evolution on islands, such as island syndromes[, ...] or the fact of explosively radiating lineages (Hughes \& Eastwood 2006, Kocher 2004)
%%, while others remain one single species (Price/Chase) (eg. \cite{price2004speciation}).
%% Also: Ecological Opportunity and Adaptive Radiation, Stroud \& Losos 2016: ``Does the concept have predictive value[...]?'' \cite{stroud2016ecological}
%% Phylogenetic niche conservatism, phylogenetic signal and the relationship between phylogenetic relatedness and ecological similarity among species, Losos 2008: ecological vs. phylogenetic similarity \cite{losos2008phylogenetic}
Still, the most mechanisms integrated in island models [?] so far include only ecological
or geo-morphological processes (Rundell \& Price 2009, Esselstyn et al. 2011),
while genetic properties are hardly recognized in the island context.
Given the essential role islands have played in the development of evolutionary theories, it is surprising that these processes
are not implemented more often in island simulations.

@@

%% Aim, objectives:
%% In this study, we aim to address...
To assess the role of genetic factors, namely genetic linkage, for island radiations
we present a mechanistic model that overcomes above-mentioned shortcomings like missing micro-evolutinary processes and lack of integrative design
by implementing genetic processes on a niche-explicit background
using an island biogeography simulation arena.
We will address the questions of what relative impact does (a) genetic linkage and
(b) reproductive isolation have on emerging species richness and composition,
and (c) which genetic and functional trait syndromes %% Which?
distinguish successful colonizer,
strong competitor and diversifying lineages.
Considering these phenomena, we expect speciation rates to be highest at intermediate levels of gene linkage --- low
levels leads to dissolving phenotype populations, whereas high levels hinder rapid adaption. %Provide Refs! %what Schiffers et al say about linkage and evolutionary rescue (basically the opposite of diversification?)
%After presenting the study questions and expectations, you should say how the questions are going to be answered  (e.g. short explain simulation experimental design).

@@

\begin{figure}
  %\includegraphics{}
  \caption{Schematic of the model's individuals genetic architecture.
    Individuals are composed of a diploid set of linkage units which combine the genes coding for functional traits.}
  \label{fig:schematic}
\end{figure}

We use an individual-based spatially and temporally explicit metacommunity model written in julia \cite{bezansonJuliaFreshApproach2017} .  \ref{fig:schematic}
The model considers competition between individuals, environmental niche adaptation, density independent mortality,
dispersal, sexual reproduction and mutation.
Individuals are characterized by a lineage identifier,
body mass, age and fitness.
Each individual's life history traits (table) are coded by mutable genes \cref{fig:schematic}.
The genes, in turn, may be combined to linkage units (``chromosomes'').
Genes in a linkage unit are considered a heredetary unit.
This means that they will always be inherited in junction.
Thus, only complete linkage units will be subject to recombination.

@@

@@

\subsection{Experimental design and scenarios}
In the simulations, we modify the value and extent of
%(a)
gene linkage
%(b) reproductive isolation (or reproduction compatibility),
%(c) habitat diversity, and
%(d) topographical complexity
to asses their relative contribution on community composition.
Additionally, we simulate lineages in isolation with varying degrees of genetic linkage
and analyize emerging species numbers.
\cref{tab:scenarios} summarizes the parameters which were varied for the scenarios, their meaning and their values. %TODO: make table.
Each parameter combination was replicated %TODO: how much?
times.%, setting a different random seed for each replicate.

@@

\begin{figure}
  %\includegraphics{}
  \caption{[2 x 2]
    Emerging numbers of species (and average population sizes) over time.
    Low-linkage lineages in COLOR1, high-linkage lineages in COLOR2.
    Top row: lineages simulated in isolation.
    !Bottom row: simulated communities.!}
  \label{specieslinkage}
\end{figure}

\begin{figure}
  %\includegraphics{}
  \caption{[2 x 2]
    ???
    Extinctions and numbers of species per lineage after XXXX timesteps against degree of genetic linkage.
    Top row: lineages simulated in isolation.
    Bottom row: simulated communities.}
  \label{specieslinkage}
\end{figure}

\begin{figure}
  %\includegraphics{}
  \caption{Trait space characteristics of lineages./Evolution of trait values over time against linkage
    (A) [Ordination] Trait space of successful colonizers compared to the mainland species pool. Each point represents a lineage.
    Arrows denote the traits.
    (B) [Ordination?] Linkage of traits of successful colonizers(/radiators?). Traits perpendicular to each other are inherited independently,
  same direction indicate tight linkage of traits.}
  \label{traitspace}
\end{figure}

@@

\section{Discussion}
%% GENETIC ARCHITECTURE/DYNAMICS STUDIED FOR DECADES. WITH GENETIC/BIG DATA, APPLY in (ISLAND) BIOGEOGRAPHY.

% Hypothesis: Species rich lineages in island systems due to adaptive radiation, thus low linkage.
% high linkage species groups maybe not even considered because different taxonomic level?

The fact that low-linkage (single) species speciated in general faster than their high-linkage counterparts
is in line with expectations given their higher potential for fast evolutionary response [as exemplified REFS].
What did come as a surprise, however was that, over time, high linkage lineages surpassed low linkage lineages
in the number of emerging species.
The reason for this is likely to be found in the high genetic variation throughout the high-linkage populations
due to genetic hitchhiking.
Combined with random drift, this genetic variation has a high potential of producing numerous species over time [non-adaptive radiation].
[EXAMPLES]
On the other hand, independent recombination of genes in low-linkage populations explains their quick adaptive response
to the conditions in their respective environments.
Once adapted and having reached a high degree of specialization, however, the genetic architecture leads to strong selection against
new and mainly deleterious mutations.
[EXAMPLES]

@@

Linkage as dynamic process: polyploidy, TEs, etc.
Here we investigate the long term effects of fixed linkage degrees on species richness.
